An Empirical Analysis of C Preprocessor Use

### 2000-01-18 macros/papers/emp-use-2.tex
Eliminate more to-do from TO-DO.

After marking the potentially expanded macros, we processed all the source
files using a program that we wrote for this study.  This program collects the
statistics about macro definitions, uses, and dependences that are
reported in the paper; more details of its operation are reported as
appropriate along with those results.  Our tool includes parsers for
expressions, statements, and declarations.  It performs approximate parsing
because the input may not be a valid C program; as a result, we may
misinterpret some constructs, but we can cope with uncompilable C and with
partial constructs in conditional compilation branches.

@@

The ten macro body categories are as follows.

@@

%% mcat_SYMBOL: 358
%%   5.3%  reserved_word (19)
%%   95%  function_name (339)
% @mcat_SYMBOL = qw( catRESERVED_WORD catFUNCTION_NAME catSYMBOLS);
\item[Symbol]
  The macro body is a single identifier that is either a function name
  (95\% of this category) or a reserved word (5\% of this category, 65\% of
  which are uses of variable names such as {\tt true} or {\tt delete} which
  are reserved words in another C dialect).  A macro body that is a macro
  name inherits that macro name's categorization rather than appearing
  here.

%% mcat_SYMBOL_UNKNOWN: 1798
%%   100%  unknown_symbol (1798)
% @mcat_SYMBOL_UNKNOWN = qw( catSYMBOL_UNKNOWN );
\item[Unknown identifier]
  The macro expands to a single identifier that is not defined in the package
  or in any library header files included by the package.  The symbol may
  be defined by compiler command arguments or may be used inside a
  conditional compilation guard because it is only meaningful 
  with a particular architecture, system, or library (for which we did not
  have header files available).

@@

The C preprocessor has capabilities outside the C programming language;
indeed, this is a primary motivation for using Cpp.  Such constructs can
present special challenges to program understanding, and especially to
reducing the use of the preprocessor by translation into C or C++.  This
section presents a list of extra-linguistic constructs whose effect is to
provide a feature unavailable in the C language.  It also reports their
frequency of appearance, both individually and in combination, in the code we analyzed.
These capabilities are based on the effects of the macro's code,
not the coarse categorization of its body; a full list of
difficult-to-understand macros would include, for example, macros classified as
syntactic as well as those described in this section.

@@

\item[side-effected formal]
        A formal argument is side-effected, as in
\begin{verbatim}
    #define SET_GLYPHS_FRAME(glyphs,frame) ((glyphs)->frame = (frame))
\end{verbatim}
        This is erroneous if the
        argument is not an lvalue\,---\,a value that can be assigned to, like
        {\tt a[i]} but unlike {\tt f(i)}.  A similar constraint applies to
        reference parameters in C++, which can model such macro arguments
        (though in C++ {\tt f(i)} can be an lvalue if {\tt f} returns a reference).
        While the compiler can catch this and some other errors, compiler
        messages can be obscure or misleading in the presence of macros,
        and our goal is to provide earlier feedback about the macro
        definition, not just some uses.

@@

An adaptation of the Indian Hill C Style and Coding
Standards~\cite{Cannon90} makes numerous suggestions. Programmers should use all capital letters for macro
names except for macros that behave like function calls, which are
acceptable only for short functions.  Statement macros should be wrapped in
{\tt do \{ \ldots{} \} while (0)}, particularly when they contain keywords.
Conditional bodies should be fully bracketed to avoid dangling semicolons.
Macros should evaluate their parameters exactly once and arguments should
not have side effects.  Side effects to macro parameters should be
documented.  Macros should avoid using global variables, since the global name may
be hidden by a local declaration.  Macros with bodies that are long or
that reference variables should take an empty parameter list.  Programmers
should use the ANSI specification rather than preprocessor tricks such as
using {\tt /**/} for token pasting and macros that rely on argument string
expansion.  Syntax should not be changed by macro substitution (except for
the {\tt PROTO} macro).  While {\tt \#ifdef} is to be avoided, it should
appear in header files (which should not be nested) rather than source
code.  It can be used to protect {\tt \#pragma} and to define macros that
are used uniformly in the code, without need for further {\tt \#ifdef}.  If
a machine is not specified, compilation should fail rather than using a
default; but conditional compilation should generally test features, not
machines or operating systems.  The text inside an {\tt \#ifdef}-bracketed section
should be parsable code, not arbitrary text.

@@

Comparing how Cpp use in libraries differs from its use in application code
may yield insights into the language needs of library authors.  Other
comparisons, such as Unix vs.~Microsoft Windows packages or packages with
different application domains or user-interface styles, may also prove valuable.

>>>>

After marking the potentially expanded macros, we processed all the source
files using a program that we wrote for this study.  This program collects the
statistics about macro definitions, uses, and dependences that are
reported in the paper; more details of its operation are reported as
appropriate along with those results.  Our tool includes approximate,
Cpp-aware parsers for
expressions, statements, and declarations.  It performs approximate parsing
because the input may not be a valid C program; as a result, we may
misinterpret some constructs, but we can cope with uncompilable C and with
partial constructs in conditional compilation branches.

@@

The ten macro body categories are as follows.  Each example in this 
{\typeofdocument} comes from the packages studied.

@@

%% mcat_SYMBOL: 358
%%   5.3%  reserved_word (19)
%%   95%  function_name (339)
% @mcat_SYMBOL = qw( catRESERVED_WORD catFUNCTION_NAME catSYMBOLS);
\item[Symbol]
  The macro body is a single identifier that is either a function name
  (95\% of this category) or a reserved word (5\% of this category, 65\% of
  which are uses of variable names such as {\tt true} or {\tt delete} which
  are reserved words in another C dialect).  Examples include
\begin{verbatim}
    #define REGEX_ALLOCATE_STACK malloc
    #define THAT this
\end{verbatim}
  A macro body that is a macro
  name inherits that macro name's categorization rather than appearing
  here.

%% mcat_SYMBOL_UNKNOWN: 1798
%%   100%  unknown_symbol (1798)
% @mcat_SYMBOL_UNKNOWN = qw( catSYMBOL_UNKNOWN );
\item[Unknown identifier]
  The macro expands to a single identifier that is not defined in the package
  or in any library header files included by the package.  For example,
\begin{verbatim}
    #define signal __bsd_signal
    #define BUFSIZE bufsize
\end{verbatim}
  The symbol may
  be defined by compiler command arguments or may be used inside a
  conditional compilation guard because it is only meaningful 
  with a particular architecture, system, or library (for which we did not
  have header files available).

@@

The C preprocessor has capabilities outside the C programming language;
indeed, this is a primary motivation for using Cpp.  Such constructs can
present special challenges to program understanding, and especially to
reducing the use of the preprocessor by translation into C or C++.  This
section presents a list of extra-linguistic constructs whose effect is to
provide a feature unavailable in the C language.  (This is orthogonal to
the ``syntactic'' category of Section~\ref{sec:categorization}, which
contains macros that act like punctuation.)  It also reports their
frequency of appearance, both individually and in combination, in the code
we analyzed.
These capabilities are based on the effects of the macro's code,
not the coarse categorization of its body; a full list of
difficult-to-understand macros would include, for example, macros classified as
syntactic as well as those described in this section.

@@

\item[side-effected formal]
        A formal argument is side-effected, as in
\begin{verbatim}
    #define POP(LOW,HIGH) do {LOW = (--top)->lo;HIGH = top->hi;} while (0)
    #define SKIP_WHITE_SPACE(p) { while (is_hor_space[*p]) p++; }
\end{verbatim}
        This is erroneous if the
        argument is not an lvalue\,---\,a value that can be assigned to, like
        {\tt a[i]} but unlike {\tt f(i)}.  A similar constraint applies to
        reference parameters in C++, which can model such macro arguments
        (though in C++ {\tt f(i)} can be an lvalue if {\tt f} returns a reference).
        While the compiler can catch this and some other errors, compiler
        messages can be obscure or misleading in the presence of macros,
        and our goal is to provide earlier feedback about the macro
        definition, not just some uses.

@@

An adaptation of the Indian Hill C Style and Coding
Standards~\cite{Cannon90} makes numerous suggestions. Programmers should use all capital letters for macro
names except for macros that behave like function calls, which are
acceptable only for short functions.  Statement macros should be wrapped in
{\tt do \verb|{| \ldots{} \verb|}| while (0)}, particularly when they contain keywords.
Conditional bodies should be fully bracketed to avoid dangling semicolons.
Macros should evaluate their parameters exactly once and arguments should
not have side effects.  Side effects to macro parameters should be
documented.  Macros should avoid using global variables, since the global name may
be hidden by a local declaration.  Macros with bodies that are long or
that reference variables should take an empty parameter list.  Programmers
should use the ANSI specification rather than preprocessor tricks such as
using {\tt /**/} for token pasting and macros that rely on argument string
expansion.  Syntax should not be changed by macro substitution (except for
the {\tt PROTO} macro).  While {\tt \#ifdef} is to be avoided, it should
appear in header files (which should not be nested) rather than source
code.  It can be used to protect {\tt \#pragma} and to define macros that
are used uniformly in the code, without need for further {\tt \#ifdef}.  If
a machine is not specified, compilation should fail rather than using a
default; but conditional compilation should generally test features, not
machines or operating systems.  The text inside an {\tt \#ifdef}-bracketed section
should be parsable code, not arbitrary text.

@@

Comparing how Cpp use in libraries differs from its use in application code
may yield insights into the language needs of library authors.  Other
comparisons, such as Unix vs.~Microsoft Windows packages, packages with
different application domains or user-interface styles, different versions
of a single package, and the like, may also prove valuable.
